XDB - An Elastic, Robust and Flexible Database Architecture for Big Data (Vision Paper)

### 2013-04-03 trunk/paper/DANAC2013/paper.tex
Applied Christophs comments to paper

Hadoop is currently the de facto standard for analyzing Big Data. Its advantages over traditional databases are its simple but flexible parallel programming model, the ability to scale to thousands of nodes as well as the robustness of its execution model. However, while highly scalable and robust, Hadoop is inherently inefficient: First, Hadoop materializes each intermediate result to achieve its high reliability, resulting in poor single query performance. Second, Hadoop uses a layered design that separates storage from execution, making data locality a substantial problem. Finally, data structures like indexes to avoid expensive scans and efficient operations such as joins are missing. 
%TODO: hard to read. Suggestion: --> Finally, data structures to avoid expensive scans, such as indexes, are missing. There is also a lack efficient implementations for operations such as joins.

This paper describes a vision of a novel elastic, robust and flexible database system called XDB and presents an initial prototype built on top of MySQL. Compared to existing approaches that focus on fixing the shortcomings of Hadoop, we take a different approach by designing XDB as middleware for an existing open-source database system to benefit from the database's maturity.
%TODO: ^^ again, reading flow feels unbalanced (at least for me). --> Compared to existing work that focuses on fixing shortcomings in Hadoop, we take a different approach by designing XDB as middleware for an existing open-source database system to benefit from the maturity of the database.
Our hypothesis is, that XDB can outperform Hadoop by at least one order of magnitude in performance while providing the same properties with regard to elasticity and robustness of the execution model as well as flexibility of the programming model. Our initial experiments presented in this paper confirm this hypothesis.

@@

\begin{itemize}
\item First, Hadoop materializes each intermediate result of a map or a reduce function either locally or into its distributed file system (HDFS). This execution model results in a poor single query performance with high resource consumption. 
\item Second, data locality is a huge problem since the storage layer is separated from the computation layer: When data is loaded in HDFS it is randomly partitioned often resulting in expensive operations, which copy data over the network.
\item Finally, efficient operations as well as index structures to read data selectively from disk are missing. Hadoop only provides unary operators and needs to read all data from disk. Thus, the first map function typically selects the relevant data by scanning all input data. Moreover, join operations are unnatural in Hadoop since binary operators are not supported leading to inefficient workarounds. 
%TODO: I believe, there should be a comma between "supported" and leading, but unsure
%TODO: (proposed partitioning aspects/sort): Also, the shuffle phase in Hadoop is designed to sort all data for partitioning, though this is not generally necessary.
\end{itemize}

@@

{\bf Objective 2:  Robust Execution-Plans}
The idea is to favor plans, whichi
%TODO: ^^ should be "plans that"
are more robust and better recoverable over potentially faster plans. In order to achieve this we use a combination of actively replicating sub-plans to make the execution more robust to node failures and stragglers (i.e., slow nodes), and by splitting the plan into independent fault-tolerant sub-plans, which materialize their intermediate query results for recovery. A cost-based optimizer has to decide how to trade robustness/fault-tolerance over execution time, i.e., when sub-plans should be replicated and when intermediate results should be materialized.  This is beneficial for long-running queries on Big Data since the all-or-nothing principle of the execution model in relational databases is replaced with a more fine-grained and more fault-tolerant execution model. \\

@@

XDB is build as a middleware for an existing database system, which is used to efficiently storing and processing the data.
%TODO: to efficiently store and process
Thus, the database layer is the lowest layer in the  XDB architecture shown in Figure \ref{fig:xdb:architecture}. The layers above form the middleware layer, which extends the database layer by the before-mentioned properties: elasticity, robustness and flexibility.

@@

\begin{figure}
\hspace{-4ex}
\epsfig{file=figures/architecture.eps,scale=0.47}
\vspace{-4ex}
\caption{XDB Architecture}
\label{fig:xdb:architecture}
\vspace{-2ex}

@@

{\bf Plan Tracker:} The plan tracker is responsible to execute and monitor multiple compile-plans. The most important component of the plan tracker is the adaptive plan executor. This component takes a compile-plan from the master tracker and divides it into partial compile-plans (sub-plans). Each sub-plan is parallelized and executed incrementally. Moreover, individual sub-plans materialize their results for robustness and adaptive parallelization.  The collection of sub-plans created for one compile-plan is called a tracker-plan in XDB. A tracker-plan can thus be seen as a data-flow graph consisting of sub-plans.

In order to execute a tracker-plan, the plan tracker uses the plan scheduler to assign compute servers to sub-plans. The plan scheduler therefore requests available compute servers from the master tracker and decides in a cost-based manner which compute servers execute which sub-plans. The main optimization goal is to avoid network traffic (i.e., optimize data locality of sub-plans). For execution, the plan tracker generates a partial execution-plan for each sub-plan and can actively deploy it on different compute servers. During code generation the plan tracker decides about the degree of parallelism once all inputs of the sub-plan are available. Thus, later sub-plans can use  exact cardinality information of their inputs. Moreover, by materializing results of sub-plans, individual sub-plans can be re-deployed on a different compute server on node-failure. The plan monitor is responsible to monitor the execution of the sub-plans and to restart them if required. \\

@@

\begin{enumerate}
\item Figure \ref{fig:xdb:exec1} shows an initial compile-plan produced by the compilation server before optimization (e.g., selections are not yet pushed down). This compile-plan is optimized in the compile server resulting in a compile-plan shown in Figure \ref{fig:xdb:exec2} (without splitting the compile-plan into sub-plans). The optimized plan shows that the selection and aggregation operators could be pushed beneath the white-box UDF operator.
\item Figure \ref{fig:xdb:exec2} additionally shows for the optimized compile-plan the sub-plans produced by the plan tracker (called tracker-plan). Each of the sub-plans is executed incrementally by the plan tracker starting with sub-plan 1 and ending with sub-plan 3.
\item For execution, each of the sub-plans in Figure \ref{fig:xdb:exec2} is parallelized and optimized individually leveraging the partitioning scheme of the database. For example, sub-plan 1 in Figure \ref{fig:xdb:exec2} is first parallelized and deployed to compute servers 1 and 2 based on the partitioning information annotated to the input tables. Assume that table \emph{lineitem} is partitioned by the attribute \emph{l\_orderkey} into two parts (i.e, part 1 and 2) such that the join in sub-plan 1 can be executed locally on each compute server. The resulting partial execution-plans for sub-plan 1 are shown by the lower two plans in Figure~\ref{fig:xdb:exec3}, which materialize their results locally into tables \emph{t1} and \emph{t2} (e.g., using memory tables in MySQL). The sub-plans 2 and 3 are executed in a similar way. Assume that the table \emph{part} is reference partitioned by the table \emph{lineitem}. In that case, the partial execution-plans for this sub-plan again can be executed locally on the compute servers 1 and 2 reading the materialized results \emph{t1} and \emph{t2}. The execution-plans for sub-plan 2 are shown in the middle of Figure~\ref{fig:xdb:exec3}). Finally, the last partial execution-plan generated for sub-plan 3 reduces the degree of parallelism to one node (i.e., compute server 1). Therefore, this partial execution-plan must read the input table~\emph{t4} remotely from compute server~2 (e.g., using a federated table in MySQL). 
\end{enumerate}

@@

In order to show the efficiency of XDB, we executed the TPC-H queries 3 and 5 on the scaling factor 30 and compared the runtime of XDB to the runtime of the Hive installation on the cluster. For XDB, we partitioned the \emph{lineitem} table into 8 parts using a hash-based partitioning scheme and all other tables using a reference based partitioning scheme such that all joins could be executed locally on each node. The results of this experiment can be seen in the left two bar groups of Figure \ref{fig:perf}.
%TODO: suggestion: "in the two leftmost bar groups"
As a result, we can see a speed-up factor of approximately 6 for both TPC-H queries.

\begin{figure}[ht] 
\hspace{-4ex}
\epsfig{file=exp/exp1.eps,scale=0.33}
\vspace{-4ex}
\caption{XDB vs. Hive (TPC-H SF=30)}
\vspace{-1ex}
\label{fig:perf}
\end{figure}

@@

@@

In this paper we presented our vision of an elastic, robust and flexible database architecture for Big Data called XDB. XDB is implemented as a middleware over an existing database system using its efficient query processing techniques. 
For execution XDB splits a compile-plan into multiple sub-plans that are optimized and executed incrementally. On the one hand this enables better optimization capabilities based on real cardinality information instead of cardinality estimates which is important when processing big data in parallel to determine an optimal degree of parallelism. On the other hand sub-plans materialize their results to allow XDB to recover from node-failures.
%TODO: ^^ these are no opposites; I would enumerate them as, e.g., "First, ..." and "Second, ..."
Moreover, the middleware supports the execution of white-box user-defined functions that can be optimized and parallelized by XDB as well to support complex analytics and other data formats in XDB as well.

>>>>

Hadoop is currently the de facto standard for analyzing Big Data. Its advantages over traditional databases are its simple but flexible parallel programming model, the ability to scale to thousands of nodes as well as the robustness of its execution model. However, while highly scalable and robust, Hadoop is inherently inefficient: First, Hadoop materializes each intermediate result to achieve its high reliability, resulting in poor single query performance. Second, Hadoop uses a layered design that separates storage from execution, making data locality a substantial problem. Finally, data structures like indexes to avoid expensive scans and efficient operations such as joins are missing.

This paper describes a vision of a novel elastic, robust and flexible database system called XDB and presents an initial prototype built on top of MySQL. Compared to existing work that focuses on fixing shortcomings in Hadoop, we take a different approach by designing XDB as middleware for an existing open-source database system to benefit from the maturity of the database.
Our hypothesis is, that XDB can outperform Hadoop by at least one order of magnitude in performance while providing the same properties with regard to elasticity and robustness of the execution model as well as flexibility of the programming model. Our initial experiments presented in this paper confirm this hypothesis.

@@

\begin{itemize}
\setlength{\itemsep}{-1pt}
\item First, Hadoop materializes each intermediate result of a map or a reduce function either locally or into its distributed file system (HDFS). This execution model results in a poor single query performance with high resource consumption. 
\item Second, data locality is a huge problem since the storage layer is separated from the computation layer: When data is loaded in HDFS it is randomly partitioned often resulting in expensive operations, which copy data over the network.
\item Third, efficient operations as well as index structures to read data selectively from disk are missing. Hadoop only provides unary operators and needs to read all data from disk. Thus, the first map function typically selects the relevant data by scanning all input data. Moreover, join operations are unnatural in Hadoop since binary operators are not supported, leading to inefficient workarounds. 
\item: FInally, Hadoop only allows a strict sequence of map and reduce operations while intermediate results are re-partitioned after each map-operation (even if it is not needed).
\end{itemize}

@@

{\bf Objective 2:  Robust Execution-Plans}
The idea is to favor plans that are more robust and better recoverable over potentially faster plans. In order to achieve this we use a combination of actively replicating sub-plans to make the execution more robust to node failures and stragglers (i.e., slow nodes), and by splitting the plan into independent fault-tolerant sub-plans, which materialize their intermediate query results for recovery. A cost-based optimizer has to decide how to trade robustness/fault-tolerance over execution time, i.e., when sub-plans should be replicated and when intermediate results should be materialized.  This is beneficial for long-running queries on Big Data since the all-or-nothing principle of the execution model in relational databases is replaced with a more fine-grained and more fault-tolerant execution model. \\

@@

XDB is build as a middleware for an existing database system, which is used to efficiently store and process the data.
Thus, the database layer is the lowest layer in the  XDB architecture shown in Figure \ref{fig:xdb:architecture}. The layers above form the middleware layer, which extends the database layer by the before-mentioned properties: elasticity, robustness and flexibility.

@@

\begin{figure}
\hspace{-4ex}
\epsfig{file=figures/architecture.eps,scale=0.47}
\vspace{-5ex}
\caption{XDB Architecture}
\label{fig:xdb:architecture}
\vspace{-3ex}

@@

{\bf Plan Tracker:} The plan tracker is responsible to execute and monitor a compile-plan while one plan tracker can monitor multiple compile-plans. The most important component of the plan tracker is the adaptive plan executor. This component takes a compile-plan from the master tracker and divides it into partial compile-plans (sub-plans). Each sub-plan is parallelized and executed individually. Moreover, individual sub-plans materialize their results for robustness and adaptive parallelization.  The collection of sub-plans created for one compile-plan is called a tracker-plan. A tracker-plan can thus be seen as a data-flow graph consisting of sub-plans. The plan tracker decides about the degree of parallelism once all inputs of a sub-plan are available. Thus, later sub-plans can use  exact cardinality information of their inputs. Moreover, by materializing results of sub-plans, individual sub-plans can be re-deployed on a different compute server on node-failure. The plan monitor of the plan tracker is responsible to monitor the execution of the sub-plans and to restart them if required.

In order to execute a tracker-plan, the plan tracker uses the plan scheduler to assign compute servers to sub-plans. The plan scheduler therefore requests compute servers from the master tracker and decides in a cost-based manner which compute servers execute which sub-plans. The main optimization goal is to avoid network traffic (i.e., optimize data locality of sub-plans). For execution, the plan tracker generates a (partial) execution-plan for each sub-plan and then deploys it on a compute server. \\

@@

\begin{enumerate}
\setlength{\itemsep}{-1pt}
\item Figure \ref{fig:xdb:exec1} shows an initial compile-plan produced by the compilation server before optimization (e.g., selections are not yet pushed down). This compile-plan is optimized in the compile server resulting in a compile-plan shown in Figure \ref{fig:xdb:exec2} (without splitting the compile-plan into sub-plans). The optimized plan shows that the selection and aggregation operators could be pushed beneath the white-box UDF operator.
\item Figure \ref{fig:xdb:exec2} additionally shows for the optimized compile-plan the sub-plans produced by the plan tracker (called tracker-plan). Each of the sub-plans is executed incrementally by the plan tracker starting with sub-plan 1 and ending with sub-plan 3.
\item For execution, each of the sub-plans in Figure \ref{fig:xdb:exec2} is parallelized and optimized individually leveraging the partitioning scheme of the database. For example, sub-plan 1 in Figure \ref{fig:xdb:exec2} is first parallelized and deployed to compute servers 1 and 2 based on the partitioning information annotated to the input tables. Assume that table \emph{lineitem} is partitioned by the attribute \emph{l\_orderkey} into two parts (i.e, part 1 and 2) such that the join in sub-plan 1 can be executed locally on each compute server. The resulting partial execution-plans for sub-plan 1 are shown by the lower two plans in Figure~\ref{fig:xdb:exec3}, which materialize their results locally into tables \emph{t1} and \emph{t2} (e.g., using memory tables in MySQL). The sub-plans 2 and 3 are executed in a similar way. Assume that the table \emph{part} is reference partitioned by the table \emph{lineitem}. In that case, the partial execution-plans for this sub-plan again can be executed locally on the compute servers 1 and 2 reading the materialized results \emph{t1} and \emph{t2}. The execution-plans for sub-plan 2 are shown in the middle of Figure~\ref{fig:xdb:exec3}). Finally, the last partial execution-plan generated for sub-plan 3 reduces the degree of parallelism to one node (i.e., compute server 1). Therefore, this partial execution-plan must read the input table~\emph{t4} remotely from compute server~2 (e.g., using a federated table in MySQL). 
\end{enumerate}

@@

In order to show the efficiency of XDB, we executed the TPC-H queries 3 and 5 on the scaling factor 30 and compared the runtime of XDB to the runtime of the Hive installation on the cluster. For XDB, we partitioned the \emph{lineitem} table into 8 parts using a hash-based partitioning scheme and all other tables using a reference based partitioning scheme such that all joins could be executed locally on each node. The results of this experiment can be seen in the two leftmost bar groups of Figure \ref{fig:perf}. As a result, we can see a speed-up factor of approximately 6 for both TPC-H queries.

\begin{figure}[ht] 
\hspace{-4ex}
\epsfig{file=exp/exp1.eps,scale=0.33}
\vspace{-3ex}
\caption{XDB vs. Hive (TPC-H SF=30)}
\vspace{-1ex}
\label{fig:perf}
\end{figure}

@@

%Dryad??

@@

In this paper we presented our vision of an elastic, robust and flexible database architecture for Big Data called XDB. XDB is implemented as a middleware over an existing database system using its efficient query processing techniques. 
For execution XDB splits a compile-plan into multiple sub-plans that are optimized and executed incrementally. First, this enables better optimization capabilities based on real cardinality information instead of cardinality estimates which is important when processing big data in parallel to determine an optimal degree of parallelism. Second, sub-plans materialize their results to allow XDB to recover from node-failures. Moreover, the middleware supports the execution of white-box user-defined functions that can be optimized and parallelized by XDB as well to support complex analytics and other data formats in XDB as well.
